---
title: "UFC_Data_Prep"
author: "Austin Snyder"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(gridExtra)
library(caret)
library(moments) 
library(reshape2)
library(knitr)
```

## Import Data
```{r}
ufc_men <- readRDS("ufc_men_cleaned.rds")
```

#Fixing rank 
- By nature, there are alot of NA's in the rank columns we will remove them. 

```{r}
rank_cols_numeric <- c("RMatchWCRank", "RWFlyweightRank", "RWFeatherweightRank", "RWStrawweightRank",
                       "RWBantamweightRank", "RHeavyweightRank", "RLightHeavyweightRank", "RMiddleweightRank",
                       "RWelterweightRank", "RLightweightRank", "RFeatherweightRank", "RBantamweightRank",
                       "RFlyweightRank", "RPFPRank", "BWFlyweightRank", "BWFeatherweightRank", "BWStrawweightRank",
                       "BWBantamweightRank", "BHeavyweightRank", "BLightHeavyweightRank", "BMiddleweightRank",
                       "BWelterweightRank", "BLightweightRank", "BFeatherweightRank", "BBantamweightRank",
                       "BFlyweightRank", "BPFPRank","BMatchWCRank")

# Remove rank-related columns
ufc_men <- ufc_men[, !(names(ufc_men) %in% rank_cols_numeric)]

```




## Transformations 
```{r}
# List your numeric features
numeric_features <- c("RedOdds", "BlueOdds", "OddsRatio", "RedExpectedValue", "BlueExpectedValue", 
                      "SigStrDif", "AvgTDDif", "KODif", "SubDif", "ReachDif", 
                      "WinDif", "LossDif", "TotalTitleBoutDif", "AgeDif")

# Melt the data
ufc_men_melted <- melt(ufc_men, measure.vars = numeric_features)

# Facet density plots
ggplot(ufc_men_melted, aes(x = value)) +
  geom_density(fill = "blue", alpha = 0.5) +
  facet_wrap(~variable, scales = "free", ncol = 4) +  # Adjust ncol to fit your layout
  labs(title = "Density Plots of Numeric Features",
       x = "",
       y = "Density") +
  theme_minimal() +
  theme(
    strip.text.x = element_text(size = 8),  # Adjust facet label size
    plot.title = element_text(hjust = 0.5)
  )
```

### Check For Skewness

```{r}

# Compute skewness for numeric columns
skew_values <- sapply(ufc_men[numeric_features], skewness, na.rm = TRUE)

# Print skewness values
print(skew_values)

```
-Odds Ratio log transformation 
- Red EV log transformation
- Blue EV log transformation
- ReachDif log transformation
- TotalTitleBoutDif log transformation

### Log Transformation

```{r}
# List of Features to Log Transform
log_transform_features <- c("OddsRatio", "RedExpectedValue", "BlueExpectedValue", 
                            "ReachDif", "TotalTitleBoutDif")

# Apply Log Transformation Correctly
for (feature in log_transform_features) {
  log_feature <- paste0("Log_", feature)  # Generate new log feature name
  
  # Ensure log transformation is applied correctly
  ufc_men[[log_feature]] <- log1p(abs(ufc_men[[feature]]))  
}

```


```{r}

# Define colors manually to keep consistency
fighter_colors <- c("Red" = "#E63946", "Blue" = "#457B9D", "neither" = "#A8DADC")

# List categorical features to plot
categorical_features <- c("Winner", "WeightClass", "BetterRank", "Finish")

# Generate plots with formatting adjustments
plot_list_cat <- lapply(categorical_features, function(feature) {
  ggplot(ufc_men, aes_string(x = feature, fill = feature)) +
    geom_bar() +
    labs(title = paste("Distribution of", feature), x = feature, y = "Count") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 30, hjust = 1, size = 10),  # Rotate x-axis labels
      axis.title = element_text(size = 12),  # Increase axis label size
      plot.title = element_text(size = 14, face = "bold")  # Increase title size
    ) +
    scale_fill_manual(values = fighter_colors)  # Apply consistent colors
})

# Arrange plots neatly
do.call(grid.arrange, c(plot_list_cat, ncol = 2))


```

- collapse decision for simplification
```{r}
ufc_men$Finish <- as.character(ufc_men$Finish)

ufc_men$Finish <- dplyr::recode(ufc_men$Finish,
                                "KO/TKO" = "KO",
                                "SUB" = "Submission",
                                "U-DEC" = "Decision",
                                "M-DEC" = "Decision",
                                "S-DEC" = "Decision",
                                .default = "UNKNOWN")  

# Convert back to factor
ufc_men$Finish <- as.factor(ufc_men$Finish)


```
- Here we combined all the decision finishes into one category. 

## Missing Values 

```{r}
colSums(is.na(ufc_men))
```

```{r}
missing_data <- data.frame(
  Feature = c("RedOdds", "BlueOdds", "RedExpectedValue", "BlueExpectedValue", 
              "BlueAvgSigStrLanded", "BlueAvgSigStrPct", "BlueAvgSubAtt", "BlueAvgTDLanded", "BlueAvgTDPct", 
              "RedAvgSigStrLanded", "RedAvgSigStrPct", "RedAvgSubAtt", "RedAvgTDLanded", "RedAvgTDPct",
              "FinishDetails", "FinishRound", "FinishRoundTime", 
              "TotalFightTimeSecs", "RedDecOdds", "BlueDecOdds", "RSubOdds", "BSubOdds", "RKOOdds", "BKOOdds", 
              "OddsRatio", "EVRatio", "LogOddsRatio", "LogEVRatio", "Log_OddsRatio", 
              "Log_RedExpectedValue", "Log_BlueExpectedValue"),
  Count = c(169, 167, 169, 167, 
            819, 680, 737, 738, 745, 
            394, 312, 312, 312, 320, 
            3087, 517, 517, 
            517, 996, 1025, 1231, 1255, 1229, 1256, 
            177, 177, 177, 177, 177, 
            169, 167),
  Percentage = round(c(169, 167, 169, 167, 
                       819, 680, 737, 738, 745, 
                       394, 312, 312, 312, 320, 
                       3087, 517, 517, 
                       517, 996, 1025, 1231, 1255, 1229, 1256, 
                       177, 177, 177, 177, 177, 
                       169, 167) / 5726 * 100, 2)  # Assuming total observations = 5726
)

# Display as a nicely formatted table
kable(missing_data, caption = "Summary of Missing Values in UFC Dataset", digits = 2)
```



### Imputation for continuous factors
- My distributions are all skewed so I will use the median to impute missing values. 

```{r}
numeric_impute_features <- c("BlueAvgSigStrLanded", "BlueAvgSigStrPct", "BlueAvgSubAtt", 
                             "BlueAvgTDLanded", "BlueAvgTDPct", "RedAvgSigStrLanded", 
                             "RedAvgSigStrPct", "RedAvgSubAtt", "RedAvgTDLanded", "RedAvgTDPct")


for (feature in numeric_impute_features) {
  ufc_men[[feature]][is.na(ufc_men[[feature]])] <- median(ufc_men[[feature]], na.rm = TRUE)
}
```


### Imputration for categorical factors

- FinishDetails is a categorical variable that has a high percentage of missing
values. NA's make up 53.92% of the data so I will remove this feature from the data set. 

-  For finish round and finish round time, I will impute the mode. 

```{r}
# Remove FinishDetails
##fc_men <- ufc_men %>% select(-FinishDetails)  # Remove the column

fill_mode <- function(x) names(sort(table(x), decreasing = TRUE))[1]

# Apply mode imputation
ufc_men$FinishRound[is.na(ufc_men$FinishRound)] <- fill_mode(ufc_men$FinishRound)
ufc_men$FinishRoundTime[is.na(ufc_men$FinishRoundTime)] <- fill_mode(ufc_men$FinishRoundTime)
```


## Imputing odds as well 

```{r}
median_impute_features <- c("RedOdds","BlueOdds", "RedExpectedValue",
                            "BlueExpectedValue","TotalFightTimeSecs", 
                             "RedDecOdds", "BlueDecOdds",
                            "RSubOdds", "BSubOdds", "RKOOdds", "BKOOdds")

for (feature in median_impute_features) {
  ufc_men[[feature]][is.na(ufc_men[[feature]])] <- median(ufc_men[[feature]], na.rm = TRUE)
}
```


```{r}
ufc_men$OddsRatio <- abs(ufc_men$RedOdds / ufc_men$BlueOdds)
ufc_men$EVRatio <- abs(ufc_men$RedExpectedValue / ufc_men$BlueExpectedValue)
```


```{r}
log_features <- c("OddsRatio", "EVRatio", "RedExpectedValue", "BlueExpectedValue")

for (feature in log_features) {
  log_feature <- paste0("Log_", feature)  # Create log version name
  ufc_men[[log_feature]] <- log1p(abs(ufc_men[[feature]]))  # Apply log transformation safely
}

```

### Remove old log features
```{r}
ufc_men <- ufc_men %>% select(-LogOddsRatio , -LogEVRatio, )
```


```{r}
colSums(is.na(ufc_men))
```

### save new data set

```{r}
## for others
write.csv(ufc_men, "ufc_men_cleaned.csv", row.names = FALSE)

## for me
saveRDS(ufc_men, "ufc_men_imputed.rds")
```


## Feaature Slection

```{r}
ufc_men_imputed <- readRDS("ufc_men_imputed.rds")
```


```{r}
# Select only numeric features
numeric_features <- names(ufc_men_imputed)[sapply(ufc_men_imputed, is.numeric)]

# Compute correlation matrix
cor_matrix <- cor(ufc_men_imputed[numeric_features], use = "complete.obs")

cor_long <- as.data.frame(as.table(cor_matrix))

# Filter for high correlations (above 0.8, but exclude self-correlations of 1)
high_correlations <- cor_long %>%
  filter(abs(Freq) > 0.8 & Var1 != Var2) %>%
  arrange(desc(abs(Freq)))

# Print only the highly correlated pairs
kable(high_correlations, caption = "Highly Correlated Features (>|0.8|)")

```

### drop higly correleated and redundant features 

- There are many highly correlated and redundant features in the data set.
- Some variables like differences used two existing variables therefore 
  will be redundant and not suitable for logistic regression. 

```{r}
features_to_drop <- c(
  "RedExpectedValue", "BlueExpectedValue", "EVRatio", "OddsRatio",
  "RedOdds", "BlueOdds", "RedWins", "RedTotalRoundsFought",
  "RedLosses", "RedLongestWinStreak", "BlueReachCms", "FinishDetails",
  "Gender", "FinishRoundTime", "Log_BlueExpectedValue",
  "TotalTitleBoutDif",  # Redundant due to BlueTotalTitleBouts
  "KODif",  # Redundant due to RedWinsByKO & BlueWinsByKO
  "SubDif",  # Redundant due to RedWinsBySubmission & BlueWinsBySubmission
  "HeightDif",  # Redundant due to RedHeightCms & BlueHeightCms
  "BlueAvgSubAtt", "RedAvgSubAtt", "AvgSubAttDif",
  "BlueWeightLbs", "RedWeightLbs",  # Redundant with WeightClass
  "Log_RedExpectedValue",  # Drop in favor of Log_EVRatio
  "BlueAvgTDLanded", "RedAvgTDLanded", "AvgTDDif",
  "BlueAvgSigStrLanded", "RedAvgSigStrLanded", "SigStrDif",
  "BlueWins", "WinDif",
  "BlueTotalRoundsFought", "TotalRoundDif", "WinStreakDif"
)

ufc_men_imputed <- ufc_men_imputed[, !(names(ufc_men_imputed) %in% features_to_drop)]
```

### Save cleaned data set
```{r}
saveRDS(ufc_men_imputed, "ufc_men_imp.rds")
```



























